---
title: "IMPUTATION TUTORIAL"
author: "Dr BASHIR SSUNA"
date: "2025-02-22"
output: html_document
---


```{r, message=FALSE, warning=FALSE}
# Importing and cleaning of datasets into R -----------------------------------------------

library(tidyverse)
library(readxl)
library(devtools)
library(zscorer)
library(dplyr)
library(factoextra)
library(knitr)
library(kableExtra)
library(gmodels)
library(ggplot2)
library(outliers)
library(tidyr)
library(VIM)
library(naniar)
library(reshape2)
library(corrplot)
library(xtable)
library(cowplot)
library(tidyr)
library(Hmisc)
library(rstudioapi)
library(broom)
library(epitools)
library(gmodels)
library(vcd)
library(clubSandwich)
library(sandwich)
library(lmtest)
library(geepack)
library(questionr)
library(mvnmle)
library(BaylorEdPsych)
library(mice)
library(lattice)
library(writexl)
# detach("package:MASS", unload = TRUE)



```




```{r}
# Clear environment 
rm(list = ls())

# set Working directory 
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))


```

Import dataset
```{r}
# Import dataset
data <- read_excel("imputation_dataset.xlsx")


## data structure
str(data)

```

#################### SELECTING VARIABLES FOR ANALYSIS ####################

```{r}
##A. Selecting categorical variables
cat_vars <- c("feature22", "feature23",
  "feature24", "feature25", "feature26", "feature27", "feature28",
  "feature29", "feature30", "feature31", "feature32"
)

##B. Selecting numeric variables
numeric_vars <- c( "feature1", "feature2", "feature3", "feature4", "feature5", 
                   "feature6", "feature7", "feature8", "feature9", "feature10", 
                   "feature11", "feature12", "feature13", "feature14", "feature15",
                   "feature16", "feature17", "feature18", "feature19", "feature20",
                   "feature21"
)

```

#################### FORMAT VARIABLES ####################

```{r}
##A. Convert categorical variables to factors
data[cat_vars] <- lapply(data[cat_vars], factor)

##B. Convert numeric variables to numeric
data[numeric_vars] <- lapply(data[numeric_vars], as.numeric)

## Check data structure
str(data)

```

############################### CHECK FOR OUTLIERS (4SD) #############################

```{r}
# General Outlier function
filter_outliers <- function(data, numeric_vars, id_var) {
  results <- list()
  
  for (var in numeric_vars) {
    var_mean <- mean(data[[var]], na.rm = TRUE)
    var_sd <- sd(data[[var]], na.rm = TRUE)
    
    lower_bound <- var_mean - 4 * var_sd
    upper_bound <- var_mean + 4 * var_sd
    
    filtered <- data %>%
      filter(data[[var]] < lower_bound | data[[var]] > upper_bound) %>%
      select(all_of(id_var), all_of(var)) %>%
      mutate(variable = var)
    
    results[[var]] <- filtered
  }
  
  final_result <- bind_rows(results)
  
  return(final_result)
}

id_var <- "ID"
outliers <- filter_outliers(data, numeric_vars, id_var)

# View the outliers
print(outliers)


```

#################### CHECK FOR MISSING VALUES ####################

```{r}
data_selected <- data %>% select(-ID, -outcome)

missing_data <- data_selected %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(Missing_Percentage = (Missing_Count / nrow(data_selected)) * 100)

print(missing_data)

```

################## NUMERIC VARIABLES THAT SHOULD NOT HAVE ZEROS ############

```{r}

non_zero_numerics <- c("feature3", "feature4", "feature8", "feature9",
                       "feature10", "feature11", "feature13", "feature14",
                       "feature15", "feature16")

### Replace 0 with NA in the specified columns
data[non_zero_numerics] <- data[non_zero_numerics] %>%
  mutate(across(all_of(non_zero_numerics), ~ replace(., . == 0, NA)))


```

#################### CHECK FOR MISSING DATA PATTERNS ####################

```{r, warning=FALSE, message=FALSE}
# Keep only columns of interest 
data_miss_patterns <- data_selected[ c(4:21)] 
data_miss_patterns %>%
  miss_var_summary()


missplot <- aggr(data_miss_patterns, col=c('turquoise', 'orange'),
                 numbers=TRUE, sortVars=TRUE, 
                 labels=names(data_miss_patterns), cex.axis=.8, 
                 gap=4, ylab=c("Missing data pattern", "Frequency"))



```


```{r}
vis_miss(data_miss_patterns)

```


```{r}
gg_miss_var(data_miss_patterns)

```


```{r}
matrixplot(data_miss_patterns)

```

#################### IMPUTE MISSING VALUES ####################

```{r, warning=FALSE, message=FALSE, results='hide'}
outcome_imputed <- mice(data, method= c(
  
  "","","","","",'pmm','pmm',"",'pmm','pmm','pmm','pmm','pmm','pmm','pmm','pmm','pmm','pmm','pmm', 'pmm','pmm','pmm','pmm', "","","","","", "","","","","", "" ), m=5, maxit=50, seed=500)

```



```{r}
# install ggmice
# devtools::install_github("amices/ggmice")
library(ggmice)

```

######### Visualize the original vs imputed dataset############

```{r}
### 1. Non imputed data (original data)

### both continuous 
ggmice(data_selected, aes(x=feature17, y = feature18))+
  geom_point()

```


```{r}
### one continuous variable
ggmice(data_selected, aes(feature1)) +
  geom_density() +
  facet_wrap(~ factor(is.na(feature5) == 0, 
                      labels = c("observed feature5", "feature5")))

```


```{r}
##### categorical variable
# ggmice(data_selected, aes(feature1)) +
#   geom_bar(fill = "white") +
#   facet_wrap(~ factor(is.na(feature28) == 0, 
#                       labels = c("observed feature28", "missing feature28")))


```

### 2. Imputed data (outcome_imputed)
 The goal here is to determine if the imputed data is similar to the original data and to choose the best imputation method

```{r}
##continuous variable
ggmice(outcome_imputed, aes( feature17, feature18)) +
  geom_point()


```


 won't show because all categorical were not missing in this dataset
```{r}
## categorical variable
ggmice(outcome_imputed, aes(feature5, feature32)) +
  geom_point()

```


```{r}
##facet wrap
ggmice(outcome_imputed, aes(feature5, feature32)) +
  geom_point() +
  facet_wrap(~ outcome, labeller = label_both)

```


```{r}
### visualize imputations (More dimension)
ggmice(outcome_imputed, aes(x = .imp, y = feature18)) +
  geom_jitter(height =  0, width = 0.25) +
  labs(x = "Imputation number")


```


```{r}
ggmice(outcome_imputed, aes(x = .imp, y = feature18)) +
  geom_jitter(height = 0, width = 0.25) +
  geom_boxplot(width = 0.5, size = 1, alpha = 0.75, outlier.shape = NA) +
  labs(x = "Imputation number")


```



```{r}
purrr::map(c("feature5", "feature18", "feature17"), ~ {
  ggmice(outcome_imputed, aes(x = .imp, y = .data[[.x]])) +
    geom_boxplot() +
    labs(x = "Imputation number")
}) %>%
  patchwork::wrap_plots()

```


```{r}
# create correlation plot
plot_corr(data_selected)

```


```{r}
# specify optional arguments
plot_corr(
  data_selected,
  vrb = c("feature1", "feature2", "feature3", "feature5", "feature17", "feature18"),
  label = TRUE,
  square = FALSE,
  diagonal = TRUE,
  rotate = TRUE
)

```


```{r}
# create predictor matrix
pred <- quickpred(data_selected)

# create predictor matrix plot
plot_pred(pred)

```


```{r}
# specify optional arguments
plot_pred(
  pred,
  label = FALSE,
  square = FALSE,
  rotate = TRUE,
  method = "pmm"
)


```


```{r}
# Generate a trace plot for a specific variable
plot(outcome_imputed, y = "feature18")

```

Lets select the first imputed dataset

```{r}

## Select dataset 1
outcome_imputed1 <- complete(outcome_imputed, 1)


```

Check if there are missing values in the imputed dataset i.e. if impuation was successful

```{r}
## Re-check for missing values
outcome_imputed1_selected <- outcome_imputed1 %>% select(-ID, -outcome)

missing_data <- outcome_imputed1_selected %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(Missing_Percentage = (Missing_Count / nrow(outcome_imputed1_selected)) * 100)

print(missing_data)

```

Save your imputed dataset to csv


```{r}
##### Write to CSV
write.csv(outcome_imputed1, "outcome_imputed.csv")

```

########################### Normality Checks ############################

```{r}
# Load necessary libraries
library(ggplot2)
library(gridExtra)  # for grid.arrange

```


```{r}
# Select continuous variables
cont_vars <- c("feature1", "feature2", "feature3", "feature4", "feature5", 
                   "feature6", "feature7", "feature8", "feature9", "feature10", 
                   "feature11", "feature12", "feature13", "feature14", "feature15",
                   "feature16", "feature17", "feature18", "feature19", "feature20",
                   "feature21")

cot_data <- outcome_imputed1[cont_vars]

```

This function will plot a histogram and a Q-Q plot for each continuous variable, along with the results of the Shapiro-Wilk normality test.

```{r}
# Function to plot histogram and normal Q-Q plot
plot_histogram_qq <- function(data, var) {
  # Histogram with density plot
  hist_plot <- ggplot(data, aes_string(x = var)) +
    geom_histogram(aes(y = ..density..), color = "black", fill = "lightblue", bins = 30) +
    geom_density(color = "red", size = 1) +
    labs(title = paste("Histogram of", var), x = var, y = "Density")
  
  # Q-Q plot
  qq_plot <- ggplot(data, aes_string(sample = var)) +
    stat_qq() +
    stat_qq_line(color = "blue") +
    labs(title = paste("Q-Q Plot of", var), x = "Theoretical Quantiles", y = "Sample Quantiles")
  
  # Arrange both plots side by side
  grid.arrange(hist_plot, qq_plot, ncol = 2)
}

# Function to perform Shapiro-Wilk normality test and plot results
normality_tests <- function(data) {
  results <- lapply(names(data), function(var) {
    if (is.numeric(data[[var]])) {
      
      # Perform Shapiro-Wilk test
      shapiro_test <- shapiro.test(data[[var]])
      
      # Display Shapiro-Wilk test result
      print(paste("Normality test for", var))
      print(shapiro_test)
      
      # Plot histogram and Q-Q plot
      plot_histogram_qq(data, var)
      
      # Return the test result as a data frame
      data.frame(variable = var,
                 W = round(shapiro_test$statistic, 3),
                 p.value = round(shapiro_test$p.value, 3))
    } else {
      NULL  
    }
  })
  
  
  test_results_df <- do.call(rbind, results)
  return(test_results_df)
}

normality_results <- normality_tests(cot_data)

# Display normality test results
print(normality_results)

```


#################### Choose between Ttest and Wilcoxon Test 

```{r}
# 
# # Ensure outcome is a factor and numeric variables are correctly referenced
cot_data <- data.frame(cot_data, outcome = outcome_imputed1$outcome)
cot_data$outcome <- as.factor(cot_data$outcome)

ttest_results <- lapply(cont_vars, function(var) {
  if (is.numeric(cot_data[[var]])) {
    # Perform the t-test
    ttest <- t.test(cot_data[[var]] ~ cot_data$outcome)

    # Calculate the group means
    mean_group1 <- mean(cot_data[[var]][cot_data$outcome == "1"],
                        na.rm = TRUE)
    SD_group1 = sd(cot_data[[var]][cot_data$outcome == "1"],
                   na.rm = TRUE)
    mean_group2 <- mean(cot_data[[var]][cot_data$outcome == "0"],
                        na.rm = TRUE)
    SD_group2 = sd(cot_data[[var]][cot_data$outcome == "0"],
                   na.rm = TRUE)

    # Calculate the mean difference
    mean_difference <- mean_group1 - mean_group2


    data.frame(
      variable = var,
      mean_group1 = round(mean_group1, 3),
      SD_group1 = round(SD_group1, 3),
      mean_group2 = round(mean_group2, 3),
      SD_group2 = round(SD_group2, 3),
      estimate = round(mean_difference, 3),
      p.value = round(ttest$p.value, 3),
      conf.low = round(ttest$conf.int[1], 3),
      conf.high = round(ttest$conf.int[2], 3)
    )
  } else {
    NULL
  }
})

ttest_results_df <- do.call(rbind, ttest_results)

# Display the results
print(ttest_results_df)


```


```{r}
### Mann-Whitney U test for Non continuous variables ####

# Ensure outcome is a factor and numeric variables are correctly referenced
# add outcome to the data
cot_data <- data.frame(cot_data, outcome = outcome_imputed1$outcome)

cot_data$outcome <- as.factor(cot_data$outcome)

mwu_results <- lapply(cont_vars, function(var) {
  if (is.numeric(cot_data[[var]])) {
    # Perform the Mann-Whitney U test
    mwu <- wilcox.test(cot_data[[var]] ~ cot_data$outcome)

    # Calculate the group medians
    median_group1 <- median(cot_data[[var]][cot_data$outcome == "1"],
                            na.rm = TRUE)
    IQR_group1 <- IQR(cot_data[[var]][cot_data$outcome == "1"],
                      na.rm = TRUE)
    median_group2 <- median(cot_data[[var]][cot_data$outcome == "0"],
                            na.rm = TRUE)
    IQR_group2 <- IQR(cot_data[[var]][cot_data$outcome == "0"],
                      na.rm = TRUE)
    # Calculate the median difference
    median_difference <- median_group1 - median_group2

    data.frame(
      variable = var,
      median_group1 = round(median_group1, 3),
      IQR_group1 = round(IQR_group1, 3),
      median_group2 = round(median_group2, 3),
      IQR_group2 = round(IQR_group2, 3),
      estimate = round(median_difference, 3),
      p.value = round(mwu$p.value, 3),
      conf.low = NA,
      conf.high = NA
    )
  } else {
    NULL
  }
})

mwu_results_df <- do.call(rbind, mwu_results)

# Display the results
print(mwu_results_df)


```

######################Categorical Variables######################

```{r}
################## Chi-square or Fishers exact  test ####################
cat_vars <- c("feature22", "feature23", "feature24", "feature25", "feature26", "feature27", "feature28",
              "feature29", "feature30", "feature31", "feature32")

cat_data <- outcome_imputed1[cat_vars]

# add outcome to the data
cat_data <- data.frame(cat_data, outcome = outcome_imputed1$outcome)

cross_table <- lapply(cat_vars, function(var) {
  if (is.factor(cat_data[[var]])) {
    # Perform the Chi-square test
    chisq <- chisq.test(cat_data[[var]], cat_data$outcome)
    
    # Perform the Fisher's exact test
    fisher <- fisher.test(cat_data[[var]], cat_data$outcome)
    
    data.frame(
      variable = var,
      chisq_statistic = round(chisq$statistic, 3),
      chisq_p.value = round(chisq$p.value, 3),
      fisher_p.value = round(fisher$p.value, 3)
    )
  } else {
    NULL
  }
})

cross_table_df <- do.call(rbind, cross_table)

# Display the results
print(cross_table_df)


# Function to perform Chi-square or Fisher's exact test
chisq.test(cat_data$feature22, cat_data$outcome)
fisher.test(cat_data$feature22, cat_data$outcome)

```

